Learning Features and Abstract Actions for Computing Generalized Plans

### 2018-11-14 doc/camera-ready/paper.tex
Another iteration by Hector

The SAT theory $T(\S,\F)$ given by formulas \eqref{eq:d1}--\eqref{eq:goal}
has $|\F| + m^2 (b^2 + 1)$ propositional variables
%has a number of propositional variables equal to $m^2 \times (b^2 + 1) + |\F|$
where $m$ is the number of expanded states in $\S$, and $b$ is their average branching factor (transitions per state).
The number of clauses is bounded by $m^2(3 + b^2 + |\F|(1+b^2))$. %$m^2 \times b^2 \times |\F|$.
%
\Omit{
This means  that  $50$ expanded states in the sample set
with an average branching factor of $10$, and a set $\F$
with $400$ candidate features, can generate up to $10^8$ clauses.
While this is a loose upper bound for the number of clauses that
follow from the implication   $D_2(s, s', t, t') \Leftharrow\ \textstyle\bigvee_f  selected(f)$
in (\label{eq:d2}) and the numbers are further reduced by symmetry consideration,
this is still a large number. ..
}
%

@@

While the initial and goal expressions $I_F$ and $G_F$ of the abstract problem can be learned 
from the satisfying assignment $\sigma$ as well, for simplicity, in the examples below they are
provided by hand.\footnote{$G_F = G_{\sigma}$ may be defined as the DNF formula whose terms 
  correspond to the abstract states over $F_{\sigma}$ that correspond to goal states in
  the sample. For each such term, a new abstract action to reach a new goal literal
  from such a term may be required.
  }
% \footnote{$G_F = G_{\sigma}$ can be defined as the DNF formula
% whose terms  correspond to the abstract states over the selected features in $F_{\sigma}$ 
% that are true in some  goal state of the sample  and false in all non-goal states.}

@@

\paragraph{Clearing a block.}
$\Q_{clear}$ contains the Blocksworld instances  with goals of the form  $clear(x)$
and stack/unstack actions. The primitive predicates, i.e., those appearing in the  instances $P$ of $\Q_{clear}$, 
are $on(\cdot,\cdot)$, $clear(\cdot)$, $ontable(\cdot)$, $holding(\cdot)$, and $handempty$. 
For this problem, a  single training instance $P$  with  5 blocks suffices to learn an abstract
model on which a general plan is computed. The  set of features $F$ learned from the theory $T_G$ is:
\begin{enumerate}[--]
  \item $H: holding$ (whether some block is being held),
  \item $X: holding \sqcap C_x$ (whether block $x$ is being held),
  \item $n(x): \abs{\exists\,on^+ . C_x}$, (number of blocks above block $x$).
\end{enumerate}

\noindent The set $A_F$  of abstract actions learned  from $T_G$ is:\footnote{Feature and action
  names are provided to make their meaning explicit; the meaning follows from their syntactic form.}

@@

The computational  bottleneck of the approach
is  in the size of the SAT  theories used to derive the features
and the actions.
% This puts a limit on the complexity of the
% feature grammars, the bound, and the number of state transitions in the sample,
% as they  result otherwise in SAT theories that are just too large (see Table~\ref{tab:exp-result-data}).
This is indeed the reason why we have chosen to use the theories $T_G$
with goal marking. Additional ideas, however, are required
to improve scalability and to make the computational model
captured by  steps 1--6 more robust. One possibility that we have not explored
is the use of incomplete (suboptimal) Max SAT solvers.

It is not fully clear whether it is reasonable in general
to assume that the features can be obtained from a large
pool of candidate features that is defined from the primitive
predicates and a general concept grammar.
%A second limitation is the assumption that 
%the features can be obtained from a large pool of candidate  features defined 
%from the primitive predicates and a  general grammar.
%It is not fully clear whether this assumption is reasonable in general.
A relevant argument made by \citeay{bonet:ijcai2018} is that generalized problems
over domains with low width \cite{nir:ecai2012}, have compact policies
in terms of general features that can be computed in polynomial time.
In addition, for dealing with arbitrary goals, the same grammar  appears
to be suitable once  \emph{goal predicates}, namely, copies $p_G$ of the primitive predicates $p$
that appear in the goal, are added to the set of primitive predicates and evaluated in the goal
\cite{martin-geffner:generalized}.

Finally, while we have dealt with instances with deterministic actions
only, the formulation can be extended to handle non-deterministic actions.
Indeed, it is sufficient to replace the formula \eqref{eq:bridge1}
that links $D1$ and $D2$ with a formula to establish that when states
$s$ and $t$ are indistinguishable (i.e.\ $\neg D1(s,t)$), then
for each action $a$ for $s$, there is an action $b$ for $t$ whose
set of transitions on $t$ is indistinguishable from the set
of transitions of $a$ on $s$; namely, for each transition $(s,a,s')$
there is a transition $(t,b,t')$ with $\Delta_f(s,s')=\Delta_f(t,t')$
for each feature $f$, and vice versa.
To do this, transitions $(s,s')$ in the sample would need to be
tagged with action labels as $(s,a,s')$.

@@

There is  an interesting  relation to the ideas
of dimensionality reduction and embeddings in machine learning
\cite{hamilton:embeddings}. The learned abstraction
maps the  states of the problem instances, 
whose size is not bounded, into valuations over a small and bounded
set of features. Moreover,  this is done while preserving the   essential properties
of states; namely,  how they are changed by actions and whether they denote  goal states or not.
The notion of \emph{sound} and \emph{complete} abstract action sets expresses this requirement
that is enforced by  the SAT formulation.

An interesting challenge for the future  is to generalize the proposed methods
to other contexts, such as learning from ``screen states'' in video games,
where the sampled states have no known structure and there are no primitive
predicates. One way for achieving this, is precisely by learning embeddings
that yield sound and complete abstractions or suitable approximations.

>>>>

The SAT theory $T(\S,\F)$ given by formulas \eqref{eq:d1}--\eqref{eq:goal}
has  $|\F| + m^2 (b^2 + 1)$ propositional variables, 
%has a number of propositional variables equal to $m^2 \times (b^2 + 1) + |\F|$
where $m$ is the number of expanded states in $\S$, and $b$ is their average branching factor (transitions per state).
The number of clauses is bounded by $m^2(2 + b^2 + |\F|(b^2+1))$. %$m^2 \times b^2 \times |\F|$.
%
\Omit{
This means  that  $50$ expanded states in the sample set
with an average branching factor of $10$, and a set $\F$
with $400$ candidate features, can generate up to $10^8$ clauses.
While this is a loose upper bound for the number of clauses that
follow from the implication   $D_2(s, s', t, t') \Leftharrow\ \textstyle\bigvee_f  selected(f)$
in (\label{eq:d2}) and the numbers are further reduced by symmetry consideration,
this is still a large number. ..
}
%

@@

While the initial and goal expressions $I_F$ and $G_F$ of the abstract problem can be learned 
from the satisfying assignment $\sigma$ as well, for simplicity, in the examples below they are
provided by hand.\footnote{$G_F = G_{\sigma}$ may be defined as the DNF formula whose terms 
  correspond to the abstract states over $F_{\sigma}$ that correspond to goal states in
  the sample.
  % For each such term, a new abstract action to reach a new goal literal
  % from such a term may be required.
  }
% \footnote{$G_F = G_{\sigma}$ can be defined as the DNF formula
% whose terms  correspond to the abstract states over the selected features in $F_{\sigma}$ 
% that are true in some  goal state of the sample  and false in all non-goal states.}

@@

\paragraph{Clearing a block.}
$\Q_{clear}$ contains the Blocksworld instances  with goals of the form  $clear(x)$
and stack/unstack actions. The primitive predicates, i.e., those appearing in the  instances $P$ of $\Q_{clear}$, 
are $on(\cdot,\cdot)$, $clear(\cdot)$, $ontable(\cdot)$, $holding(\cdot)$, and $handempty$. 
For this problem, a  single training instance $P$  with  5 blocks suffices to learn an abstract
model from  which a general plan is computed. The  set of features $F$ learned from the theory $T_G$ is:
\begin{enumerate}[--]
  \item $H: holding$ (whether some block is being held),
  \item $X: holding \sqcap C_x$ (whether block $x$ is being held),
  \item $n(x): \abs{\exists\,on^+ . C_x}$, (number of blocks above block $x$).
\end{enumerate}

\noindent The set $A_F$  of abstract actions learned is:\footnote{Feature and action
  names are provided to make their meaning explicit; the meaning follows from their syntactic form.}

@@

The computational  bottleneck of the approach
is  in the size of the SAT  theories used to derive the features
and the actions.
% This puts a limit on the complexity of the
% feature grammars, the bound, and the number of state transitions in the sample,
% as they  result otherwise in SAT theories that are just too large (see Table~\ref{tab:exp-result-data}).
This is the   reason why we have chosen to use the more compact theories $T_G$
in the experiments.  Additional ideas, however, are required
to improve scalability and to make the computational model
captured by  steps 1--6 more robust.
% One possibility that we have not explored
% is the use of incomplete (suboptimal) Max SAT solvers.

From the point of view of expressivity, it is not clear how restrictive
is the assumption  that the  features can be obtained from a pool of  features
defined from the primitive predicates and a general grammar.
%A second limitation is the assumption that 
%the features can be obtained from a large pool of candidate  features defined 
%from the primitive predicates and a  general grammar.
%It is not fully clear whether this assumption is reasonable in general.
A relevant argument made by \citeay{bonet:ijcai2018} is that generalized problems
over domains with low width \cite{nir:ecai2012}, have compact policies
in terms of features that can be computed in polynomial time.
It is an open question, however,  whether such features are  captured
by the proposed grammar or a  suitable variation. The feature language, however, seems 
adequate for  dealing with arbitrary goals, once  \emph{goal predicates} are added
to the set of primitive predicates. Goal predicates are copies $p_G$ of the primitive predicates $p$ that
appear in the goal and which are evaluated in the goal  \cite{martin-geffner:generalized}.

Finally, the formulation can be extended to handle non-deterministic actions.
For this, it is sufficient to replace the formula \eqref{eq:bridge1}
that links $D_1$ and $D_2$ atoms  with a formula asserting that, 
if two states $s$ and $t$ are indistinguishable (i.e.\ $\neg D_1(s,t)$), then
for each action $a$ in $s$, there must be  an action $b$ in  $t$,
such that the  state transitions $(t,b,t')$ generated by $b$ in $t$, 
can't be distinguished from the transitions $(s,a,s')$ generated by $a$ in $s$. 
For this, transitions $(s,s')$ in the sample would need to be tagged with the actions
that generated them as  $(s,a,s')$.

@@

There is  an interesting  relation between sound and complete abstractions, on the one hand, 
and the ideas of dimensionality reduction and embeddings in machine learning, on the other 
\cite{hamilton:embeddings}. Sound and complete abstractions  map  the  states of the problem
instances,  whose size is not bounded, into valuations over a small and bounded set of features,
while preserving the   essential properties of states; namely,  how they are changed by actions
and whether they denote  goal states or not.
% The notion of \emph{sound} and \emph{complete} abstract action sets expresses this requirement
% that is enforced by  the SAT formulation.
An interesting challenge for the future  is to generalize the proposed methods
to other contexts, such as learning from ``screen states'' in video games,
where the sampled states have no known structure and there are no primitive
predicates. One way for achieving this would be precisely by learning embeddings
that yield sound and complete abstractions or suitable approximations.
